---
title: "Project Phase 2"
author: "Alex Song"
date: "2026-03-07"
output:
  html_document: default
  word_document: default
---

- instant: record index
- dteday : date
c - season : season (1:springer, 2:summer, 3:fall, 4:winter)
c - yr : year (0: 2011, 1:2012)
c - mnth : month ( 1 to 12)
c - holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
c - weekday : day of the week
c - workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
c - weathersit : 
    - 1: Clear, Few clouds, Partly cloudy, Partly cloudy
    - 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
    - 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
    - 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog

- temp : Normalized temperature in Celsius. The values are divided to 41 (max)
- atemp: Normalized feeling temperature in Celsius. The values are divided to 50 (max)
- hum: Normalized humidity. The values are divided to 100 (max)
- windspeed: Normalized wind speed. The values are divided to 67 (max)
- casual: count of casual users
- registered: count of registered users
- cnt: count of total rental bikes including both casual and registered

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(glmnet)
library(pls)
library(mgcv)
```

### Loading and preparing the data
```{r}
set.seed(123)

bike <- read.csv("C:/Users/alexm/Desktop/DATA 5321/Project/bike_sharing_daily.csv")
#head(bike)

# Reset Bike Data
# Training - 2011 (yr = 0), Test - 2012 (yr = 1)
bike_2011 <- bike %>%
  filter(yr == 0) %>%
  select(-instant, -yr)

bike_2012 <- bike %>%
  filter(yr == 1) %>%
  select(-instant, -yr)

# Factorize categorical variables
cat_vars <- c("season", "mnth", "holiday", "weekday", "workingday", "weathersit")

bike_2011 <- bike_2011 %>%
  mutate(across(all_of(cat_vars), as.factor))

bike_2012 <- bike_2012 %>%
  mutate(across(all_of(cat_vars), as.factor))

```

# Phase 2
```{r}
# Create Model Matrices for Ridge/Lasso
x <- model.matrix(cnt ~ . - casual - registered - cnt - dteday, data = bike_2011)[, -1]
y <- bike_2011$cnt

x_test <- model.matrix(cnt ~ . - casual - registered - cnt - dteday, data = bike_2012)[, -1]
y_test <- bike_2012$cnt

# Create Dataframes for PCR/PLS
bike_2011_clean <- bike_2011 %>% select(-casual, -registered, -dteday)
bike_2012_clean <- bike_2012 %>% select(-casual, -registered, -dteday)
```

## Phase 2 - Part 1a: Lasso & Ridge
```{r}
set.seed(123)

# Fit Models with CV
cv_ridge <- cv.glmnet(x, y, alpha = 0)
cv_lasso <- cv.glmnet(x, y, alpha = 1)

# Plot CV Curves for Optimal Lambda
par(mfrow = c(1, 2))
plot(cv_ridge)
title("Ridge Cross-Validation Curve", line = 2.5)
plot(cv_lasso)
title("Lasso Cross-Validation Curve", line = 2.5)
par(mfrow = c(1, 1))
```

```{r}
# Print Optimal Lambdas
cat("\n--- Optimal Lambdas ---\n")
cat("Ridge Lambda Min:", cv_ridge$lambda.min, "\n")
cat("Lasso Lambda Min:", cv_lasso$lambda.min, "\n")

# Print Coefficients at Optimal Lambda
cat("\n--- Ridge Coefficients (lambda.min) ---\n")
print(coef(cv_ridge, s = "lambda.min"))
cat("\n--- Lasso Coefficients (lambda.min) ---\n")
print(coef(cv_lasso, s = "lambda.min"))
```

```{r}
# Calculate Training RMSE
pred_train_ridge <- predict(cv_ridge, newx = x, s = "lambda.min")
pred_train_lasso <- predict(cv_lasso, newx = x, s = "lambda.min")

cat("\n--- Training Set RMSE ---\n")
cat("Ridge:", sqrt(mean((y - pred_train_ridge)^2)), "\n")
cat("Lasso:", sqrt(mean((y - pred_train_lasso)^2)), "\n")
```


## Phase 2 - Part 1b: PCR and PLS
```{r}
set.seed(123)

# Fit Models
pcr_model <- pcr(cnt ~ ., data = bike_2011_clean, scale = TRUE, validation = "CV")
pls_model <- plsr(cnt ~ ., data = bike_2011_clean, scale = TRUE, validation = "CV")

# Plot RMSEP Curves for Component Selection
par(mfrow = c(1, 2))
plot(RMSEP(pcr_model), main = "PCR RMSEP")
plot(RMSEP(pls_model), main = "PLS RMSEP")
par(mfrow = c(1, 1))
```

```{r}
# Print Summary
cat("\n--- PCR Model Summary ---\n")
summary(pcr_model)
cat("\n--- PLS Model Summary ---\n")
summary(pls_model)
```

```{r}
# Optimal Components
pcr_ncomp_opt <- which.min(RMSEP(pcr_model)$val[1,1,]) - 1
pls_ncomp_opt <- which.min(RMSEP(pls_model)$val[1,1,]) - 1

cat("\n--- Optimal Components ---\n")
cat("PCR:", pcr_ncomp_opt, "\n")
cat("PLS:", pls_ncomp_opt, "\n")
```

```{r}
# Training RMSE
pred_pcr_train <- predict(pcr_model, bike_2011_clean, ncomp = pcr_ncomp_opt)
pred_pls_train <- predict(pls_model, bike_2011_clean, ncomp = pls_ncomp_opt)

cat("\n--- Training Set RMSE ---\n")
cat("PCR:", sqrt(mean((bike_2011_clean$cnt - pred_pcr_train)^2)), "\n")
cat("PLS:", sqrt(mean((bike_2011_clean$cnt - pred_pls_train)^2)), "\n")
```

## Phase 2 - Part 2: GAM
```{r}
# Fit GAM with splines for temp, hum, windspeed
gam_model <- gam(
  cnt ~ season + mnth + holiday + weekday + workingday + weathersit +
    s(temp, bs = "cs") + 
    s(hum, bs = "cs") + 
    s(windspeed, bs = "cs"),
  data = bike_2011
)

# Print Summary
cat("\n--- GAM Model Summary ---\n")
summary(gam_model)
```

```{r}
# Plot Partial Effects (all)
plot(gam_model, pages = 1, all.terms = TRUE, main = "Gam Partial Effects")

```

```{r}
# Plots 3 Spline Terms
plot.gam(gam_model, select = 1, col = "blue", main = "Effect of Temperature")
plot.gam(gam_model, select = 2, col = "darkgreen", main = "Effect of Humidity")
plot.gam(gam_model, select = 3, col = "red", main = "Effect of Windspeed")
layout(1)  # reset

```

```{r}
# Training RMSE
pred_gam_train <- predict(gam_model, newdata = bike_2011)
train_rmse <- sqrt(mean((bike_2011$cnt - pred_gam_train)^2))

cat("\n--- GAM Training Set RMSE ---\n")
cat("GAM:", train_rmse, "\n")
```

## Phase 2 - Part 3: Test Set Evaluation and Final Model Selection
```{r}
calc_mae <- function(actual, predicted) { mean(abs(actual - predicted)) }

# Predictions
pred_ridge_test <- predict(cv_ridge, newx = x_test, s = "lambda.min")
pred_lasso_test <- predict(cv_lasso, newx = x_test, s = "lambda.min")
pred_pcr_test <- as.vector(predict(pcr_model, newdata = bike_2012_clean, ncomp = pcr_ncomp_opt))
pred_pls_test <- as.vector(predict(pls_model, newdata = bike_2012_clean, ncomp = pls_ncomp_opt))
pred_gam_test <- predict(gam_model, newdata = bike_2012)

# Compile Results
results <- data.frame(
  Model = c("Ridge", "Lasso", "PCR", "PLS", "GAM"),
  RMSE = c(
    sqrt(mean((y_test - pred_ridge_test)^2)),
    sqrt(mean((y_test - pred_lasso_test)^2)),
    sqrt(mean((bike_2012_clean$cnt - pred_pcr_test)^2)),
    sqrt(mean((bike_2012_clean$cnt - pred_pls_test)^2)),
    sqrt(mean((bike_2012_clean$cnt - pred_gam_test)^2))
  ),
  MAE = c(
    calc_mae(y_test, pred_ridge_test),
    calc_mae(y_test, pred_lasso_test),
    calc_mae(bike_2012_clean$cnt, pred_pcr_test),
    calc_mae(bike_2012_clean$cnt, pred_pls_test),
    calc_mae(bike_2012_clean$cnt, pred_gam_test)
  )
)

cat("\n--- Final Test Set Performance (2012) ---\n")
print(results)

# Print Best Model
best_model_row <- results[which.min(results$RMSE), ]
cat("\n--- BEST MODEL SELECTED ---\n")
print(best_model_row)
```

```{r}
# --- Training Performance (2011) ---
# Predictions on training data
pred_gam_train <- predict(gam_model, newdata = bike_2011)

# Residuals
resid_train <- bike_2011$cnt - pred_gam_train

# Training metrics
rmse_2011 <- sqrt(mean(resid_train^2))
mae_2011  <- mean(abs(resid_train))

cat("\n--- GAM Training Performance (2011) ---\n")
cat("RMSE:", rmse_2011, "\n")
cat("MAE :", mae_2011,  "\n")


# --- Test Performance (2012) ---
# Predictions on test data
pred_gam_test <- predict(gam_model, newdata = bike_2012)

# Residuals
resid_test <- bike_2012$cnt - pred_gam_test

# Test metrics
rmse_2012 <- sqrt(mean(resid_test^2))
mae_2012  <- mean(abs(resid_test))

cat("\n--- GAM Test Performance (2012) ---\n")
cat("RMSE:", rmse_2012, "\n")
cat("MAE :", mae_2012,  "\n")
```

```{r}
# Observed vs Predicted Plot for GAM
plot(
  bike_2012$cnt, pred_gam_test,
  xlab = "Observed Count",
  ylab = "Predicted Count",
  main = "Observed vs Predicted: GAM",
  pch = 19, col = rgb(0, 0, 1, 0.4)
)
abline(0, 1, col = "red", lwd = 2)
```


